Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-03-28 parts/ccl.tex
global proofread

Further work can be derived from what have been presented in this paper.
The over-approximation on Process Hitting models without priorities proposed in~\cite{PMR12-MSCS}
is still accurate in the framework with priorities (by “flattening” all actions),
but may be refined given the restrictions proposed in this paper,
and a specific search of key processes or cut sets may be derived.
%but turns out to be too wide even in some obvious cases that are consequently not conclusive.
%This approximation may be refined in order to better fit the introduction of priorities, and mak the overall approximation approach 
%and mode precisely the class of models studied in this paper.
Furthermore, we are investigating alternative under-approximations that can be
applied directly on a larger class of Process Hitting models, \ie{} without an
intermediate encoding; our objective being to increase the conclusiveness of the static
analysis approach.
Finally, in order to take into account quantitative data in transition delays, the overall approximation method could be extended to handle evolutions that are chronometric instead of only chronologic.

>>>>

Further work can be derived from what have been presented in this paper.
The over-approximation on Process Hitting models without priorities proposed in~\cite{PMR12-MSCS}
is still accurate in the framework with priorities (by “merging” all actions),
but may be refined given the restrictions proposed in this paper,
and a specific search of key processes or cut sets may be derived.
%but turns out to be too wide even in some obvious cases that are consequently not conclusive.
%This approximation may be refined in order to better fit the introduction of priorities, and mak the overall approximation approach 
%and mode precisely the class of models studied in this paper.
Furthermore, we are investigating alternative under-approximations that can be
applied directly on a larger class of Process Hitting models, \ie without an
intermediate encoding; our objective being to increase the conclusiveness of the static
analysis approach.
Finally, in order to take into account quantitative data in transition delays, the overall approximation method could be extended to handle evolutions that are chronometric instead of only chronologic.

### 2014-03-28 parts/def-ph.tex
global proofread

\section{The Process Hitting Framework}

@@

\subsection{Definition of the Process Hitting with $k$ classes of priorities}
\label{ssec:PH}
A PH with $k$ classes of priorities (\pref{def:ph}), also simply called “PH” in the following when it is not ambiguous, gathers a finite number of concurrent \emph{processes} divided into a finite set of \emph{sorts}.
A process belongs to a unique sort and is noted $a_i$ where $a$ is the sort and $i$ the identifier of the process within the sort $a$.
Each process stands for a kind of “activity level” of its sort; a state of the PH thus corresponds to a set of processes containing exactly one process of each sort.

@@

\begin{definition}[Process Hitting with $k$ classes of priorities]
\label{def:ph}
  If $k \in \sNN$, a \emph{Process Hitting with $k$ classes of priorities} is a triplet $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$,
  where $\PHa^{\langle k \rangle} = (\PHa^{(1)}; \dots; \PHa^{(k)})$ is a $k$-tuple and:
  \begin{itemize}
    \item $\PHs \DEF \{a, b, \dots\}$ is the finite set of \emph{sorts},
    \item $\PHl \DEF \underset{a \in \PHs}{\times} \PHl_a$ is the finite set of \emph{states}, where $\PHl_a = \{a_0, \ldots, a_{l_a}\}$ is the finite set of \emph{processes} of sort $a \in \PHs$ and $l_a \in \sN^*$. Each process belongs to a unique sort: $\forall (a_i; b_j) \in \PHl_a \times \PHl_b, a \neq b \Rightarrow a_i \neq b_j$,
    \item $\forall n \in \llbracket 1; k \rrbracket, \PHa^{(n)} \DEF \{\PHfrappe{a_i}{b_j}{b_l} \mid (a; b) \in \PHs^2 \wedge (a_i; b_j; b_l) \in \PHl_a \times \PHl_b \times \PHl_b \wedge b_j \neq b_l \wedge a = b \Rightarrow a_i = b_j \}$ is the finite set of \emph{actions of priority $n$}.
  \end{itemize}
  We call $\PHproc \DEF \bigcup_{a \in \PHs} \PHl_a$ the set of all processes, and $\PHh \DEF \bigcup_{n \in \segm{1}{k}} \PHh^{(n)}$ the set of all actions.
\end{definition}
\noindent
The sort of a process $a_i$ is referred to as $\PHsort(a_i) = a$.
Given a state $s\in \PHl$, the process of sort $a \in \PHs$ present in $s$ is denoted by $\PHget{s}{a}$, that is the $a$-coordinate of the state $s$.
If $a_i \in \PHl_a$, we define the notation $a_i \in s \EQDEF \PHget{s}{a} = a_i$.
The override of a state $s$ by a process $a_i$ is defined in \pref{def:statecap} as the same state in which the process of sort $a$ has been replace by $a_i$,
which then allows to define the dynamics of a PH in \pref{def:play}.
\begin{definition}[$\Cap : \PHl \times \PHproc \rightarrow \PHl$]
\label{def:statecap}
  Given a state $s \in \PHl$ and a process $a_i \in \PHproc$, $(s \Cap a_i)$ is the state defined by:
  $\PHget{(s \Cap a_i)}{a} = a_i \wedge \forall b \neq a, \PHget{(s \Cap a_i)}{b} = \PHget{s}{b}$.
  We also extend this definition to a set of processes $ps$ given that all processes are from different sorts by the override of each process:
  $\forall as \subseteq \PHs, \forall ps \in \underset{a \in as}{\times} \PHl_a, s \Cap ps = s \underset{a_i \in ps}{\Cap} a_i$.
\end{definition}
\begin{definition}[Dynamics of a PH ($\PHPtrans$)]
\label{def:play}
  An action $h = \PHhit{a_i}{b_j}{b_k} \in \PHa^{(n)}$ of priority $n$ is \emph{playable} in $s \in \PHl$
  if and only if $\PHget{s}{a} = a_i$, $\PHget{s}{b} = b_j$ and $\forall m < n, \forall g \in \PHa^{(m)}, \PHhitter(g) \notin s \vee \PHtarget(g) \notin s$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play of the action $h$ in $s$ and is defined by: $(s \PHplay h) = s \Cap \PHbounce(h)$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.

@@

\pref{def:substate} defines the notion of sub-state on a set of sorts,
that is used to consider the interesting part of a complete state.
\begin{definition}[Sub-states ($\PHsublize{\PHl}$)]
\label{def:substate}
  If $S \subset \PHs$ is a set of sorts, a sub-set on $S$ is an element of:
  $\PHsubl[\PHl]_S \DEF \bigtimes{a \in S} \PHl_a$.
  The set of all sub-sets is:
  $\PHsubl[\PHl] \DEF \bigcup_{S \in\powerset(\PHs)} \PHsubl[\PHl]_S$.

  \noindent
  Furthermore, if $\mysigma \in \PHsubl[\PHl]$ and $s \in \PHl$, we note:
    \[\mysigma \subseteq s \EQDEF \forall a_i \in \Proc, a_i \in \mysigma \Rightarrow a_i \in s \enspace.\]
\end{definition}
We note that a state is \textit{a fortiori} a sub-state: $\PHl \subset \PHsubl[\PHl]$.

@@

	\path[bounce, bend right=55]
      \TBounce{ab_0}{}{ab_2}{.west}
      \TBounce{ab_1}{}{ab_3}{.west}
	;
	\path[bounce, bend left=20]
      \TBounce{ab_3}{}{ab_1}{.south east}
      \TBounce{ab_2}{}{ab_0}{.south east}
	;
    \path[bounce, bend right=20]
      \TBounce{ab_3}{}{ab_2}{.north east}
      \TBounce{ab_1}{}{ab_0}{.north east}
    ;
    \path[bounce, bend left=30]
      \TBounce{ab_2}{}{ab_3}{.west}
      \TBounce{ab_0}{}{ab_1}{.west}
    ;
    \path[bounce, bend right]
      \TBounce{a_1}{}{a_0}{.north west}
      \TBounce{b_1}{}{b_0}{.north west}
    ;
    \path[bounce, bend left]
      \TBounce{a_0}{}{a_1}{.south west}
      \TBounce{b_0}{}{b_1}{.south west}
    ;
    \path[bounce, bend left]
      \TBounce{c_0}{}{c_1}{.south west}
    ;
    \TState{a_1, b_0, ab_2, c_0}
  \end{tikzpicture}
  }
  \caption{
  \label{fig:ph-livelock}
    An example of PH with $2$ classes of priorities.
    Sorts are represented as labelled boxes and processes as circles with their identifier on the side.
    Actions of $\PHh^{(1)}$ are represented by thick arrows and actions of $\PHh^{(2)}$ are represented by single arrows;
    the hit part of each action in drawn in plain line and the bounce part is in dotted line.
    Greyed processes stand for a possible state $s = \PHstate{a_1, b_0, c_0, ab_{10}}$.
  }
\end{figure}

@@

The construction of cooperation in PH allows to encode any Boolean function between cooperating processes \cite{PMR10-TCSB}.
Due to the introduction of priorities into the PH framework,
it is possible to build cooperations with no temporal shift by defining actions updating the cooperative sorts with the highest class of priority.
This allows to gain the same expressivity in PH than in Boolean networks, as stated in \pref{ssec:hypothesis}.
The aim of this paper is to allow the static analysis of the dynamics to be handled on PH models comprising such higher priority actions updating cooperative sorts.

@@

In the scope of this paper, we focus on a specific class of PH models
in which actions of $\PHh^{(1)}$ are called \emph{prioritized actions} and allow to model non-biological actions.
%and show that they are equivalent to discrete networks.
We consider in this section a PH model with $k$ classes of priorities: $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$, with $k \in \sNN$
and we define here the restrictions that lead to this class of models.

%\pref{cr:2prio} allows to distinguish two kinds of actions:
%\emph{unprioritised actions} modelling the non-determinacy of biological evolutions
%and \emph{prioritised actions} used to model non-biological behaviours in the model, namely the update of cooperative sorts.
\pref{cr:bounded} states that the dynamics of the studied model $\PH$ contains no infinite sequence of prioritised actions.
As these actions can be considered as non-biological and therefore instantaneous, we thus prevent the existence of any Zeno-like behaviour
which would allow the play of an infinite sequence of prioritised actions in “zero time”.
As a consequence, the set $\restriction{\Sce}{1}$ is finite.
%\begin{condition}[2 classes of priorities]
%\label{cr:2prio}
%  In this paper, we only consider Process Hitting with $2$ classes of priorities:
%  $\PH = (\PHs; \PHl; \PHa^{\langle 2 \rangle})$.
%\end{condition}
%
\begin{condition}[Bounded termination]
\label{cr:bounded}
  The dynamics of $\restriction{\PH}{1}$ contains no cycles:
  $\exists N \in \sN, \forall s \in \PHl, \forall \delta \in \restriction{\Sce}{1}(s), |\delta| \leq N$.
\end{condition}

@@

Our second condition (\pref{cr:compcs}) imposes that any sort in the Process
Hitting is either a well-formed component (\pref{def:component}) or a well-formed cooperative sort
(\pref{def:cs}).
The former is a sort that is hit only by actions with priority greater than $2$.
The latter is a sort that is hit only by actions with priority $1$ and which
always converge to the same process with respect to the state of the components
that have an influence on it (\pref{def:conn}).
We note $\components\subset\Sigma$ the set of well-formed components,
and $\cs\subset\Sigma$ the set of well-formed cooperative sorts.

@@

\begin{definition}[Components influence $(\compin: \PHs\to\powerset(\components))$]
\label{def:conn}
Given a sort $a$, $\compin(a)\DEF \conn(a)\cap\components$ where
$\conn(a)$ is the smallest set of sorts satisfying the following properties:
\begin{align*}
a\in\conn(a) &
\\
\forall h\in  \PHh^{(1)},
	\Sigma(\target{h})\in\conn(a) & \Longrightarrow \Sigma(\hitter{h})\in\conn(a)
\end{align*}
\end{definition}

@@

@@

\begin{example}
  The PH in \pref{fig:ph-livelock} contains three components $a$, $b$ and $c$ and a cooperative sort $ab$ that models cooperation between sorts $a$ and $b$.
\end{example}

>>>>

\section{The Process Hitting Framework}
\label{sec:ph}

@@

\subsection{Definition of the Process Hitting with $k$ classes of priorities}
\label{ssec:PH}
A PH with $k \in \sN^*$ classes of priorities (\pref{def:ph}), also simply called “PH” in the following when it is not ambiguous, gathers a finite number of concurrent \emph{processes} divided into a finite set of \emph{sorts}.
A process belongs to a unique sort and is noted $a_i$ where $a$ is the sort and $i$ the identifier of the process within the sort $a$.
Each process stands for a kind of “activity level” of its sort; a state of the PH thus corresponds to a set of processes containing exactly one process of each sort.

@@

\begin{definition}[Process Hitting with $k$ classes of priorities]
\label{def:ph}
  If $k \in \sN^*$, a \emph{Process Hitting with $k$ classes of priorities} is a triplet $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$,
  where $\PHa^{\langle k \rangle} = (\PHa^{(1)}; \dots; \PHa^{(k)})$ is a $k$-tuple and:
  \begin{itemize}
    \item $\PHs \DEF \{a, b, \dots\}$ is the finite set of \emph{sorts},
    \item $\PHl \DEF \underset{a \in \PHs}{\times} \PHl_a$ is the finite set of \emph{states}, where $\PHl_a = \{a_0, \ldots, a_{l_a}\}$ is the finite set of \emph{processes} of sort $a \in \PHs$ and $l_a \in \sN^*$. Each process belongs to a unique sort: $\forall (a_i; b_j) \in \PHl_a \times \PHl_b, a \neq b \Rightarrow a_i \neq b_j$,
    \item $\forall n \in \llbracket 1; k \rrbracket, \PHa^{(n)} \DEF \{\PHfrappe{a_i}{b_j}{b_l} \mid (a; b) \in \PHs^2 \wedge (a_i; b_j; b_l) \in \PHl_a \times \PHl_b \times \PHl_b \wedge b_j \neq b_l \wedge a = b \Rightarrow a_i = b_j \}$ is the finite set of \emph{actions of priority $n$}.
  \end{itemize}
  We call $\PHproc \DEF \bigcup_{a \in \PHs} \PHl_a$ the set of all processes, and $\PHh \DEF \bigcup_{n \in \segm{1}{k}} \PHh^{(n)}$ the set of all actions.
  Furthermore, for all $n \in \sN^*$ and $h \in \PHh^{(n)}$, $\prio(h) \DEF n$.
\end{definition}
%
\noindent
The sort of a process $a_i$ is referred to as $\PHsort(a_i) = a$.
Given a state $s\in \PHl$, the process of sort $a \in \PHs$ present in $s$ is denoted by $\PHget{s}{a}$, that is the $a$-coordinate of the state $s$.
If $a_i \in \PHl_a$, we define the notation $a_i \in s \EQDEF \PHget{s}{a} = a_i$.
\pref{def:substate} defines the notion of sub-state on a set of sorts, that is used to focus on the interesting part of a complete state.
Note that a state is \textit{a fortiori} a sub-state.
The override of a state $s$ by a process $a_i$ (or by a set of processes)
is defined in \pref{def:statecap} as the same state in which the process of sort $a$ has been replaced by $a_i$,
which then allows to define the dynamics of a PH in \pref{def:play}.
%
\begin{definition}[Sub-states ($\PHsublize{\PHl}$)]
\label{def:substate}
  If $S \subset \PHs$ is a set of sorts, a sub-set on $S$ is an element of:
  $\PHsubl[\PHl]_S \DEF \bigtimes{a \in S} \PHl_a$.
  The set of all sub-sets is:
  $\PHsubl[\PHl] \DEF \bigcup_{S \in\powerset(\PHs)} \PHsubl[\PHl]_S$.

  \noindent
  Furthermore, if $\mysigma \in \PHsubl[\PHl]$ and $s \in \PHl$, we note:
    \[\mysigma \subseteq s \EQDEF \forall a_i \in \Proc, a_i \in \mysigma \Rightarrow a_i \in s \enspace.\]
\end{definition}
%
\begin{definition}[$\Cap : \PHl \times \PHproc \rightarrow \PHl$]
\label{def:statecap}
  Given a state $s \in \PHl$ and a process $a_i \in \PHproc$, $(s \Cap a_i)$ is the state defined by:
  $\PHget{(s \Cap a_i)}{a} = a_i \wedge \forall b \neq a, \PHget{(s \Cap a_i)}{b} = \PHget{s}{b}$.
  We also extend this definition to a set of processes given that all processes are from different sorts by the override of each process:
  $\forall ps \in \PHsubl[\PHl], s \Cap \toset{ps} = s \underset{a_i \in \toset{ps}}{\Cap} a_i$.
  %$\forall as \subseteq \PHs, \forall ps \in \underset{a \in as}{\times} \PHl_a, s \Cap ps = s \underset{a_i \in ps}{\Cap} a_i$.
\end{definition}
%
\begin{definition}[Dynamics of a PH ($\PHPtrans$)]
\label{def:play}
  An action $h = \PHhit{a_i}{b_j}{b_k} \in \PHa^{(n)}$ of priority $n$ is \emph{playable} in $s \in \PHl$
  if and only if $\PHget{s}{a} = a_i$, $\PHget{s}{b} = b_j$ and $\forall m < n, \forall g \in \PHa^{(m)}, \PHhitter(g) \notin s \vee \PHtarget(g) \notin s$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play of the action $h$ in $s$ and is defined by: $(s \PHplay h) = s \Cap b_k$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.

@@

@@

	\path[bounce, bend right=55]
      \TBounce{ab_0}{}{ab_2}{.west}
      \TBounce{ab_1}{}{ab_3}{.west}
	;
	\path[bounce, bend left=20]
      \TBounce{ab_3}{}{ab_1}{.south east}
      \TBounce{ab_2}{}{ab_0}{.south east}
	;
    \path[bounce, bend right=20]
      \TBounce{ab_3}{}{ab_2}{.north east}
      \TBounce{ab_1}{}{ab_0}{.north east}
    ;
    \path[bounce, bend left=30]
      \TBounce{ab_2}{}{ab_3}{.west}
      \TBounce{ab_0}{}{ab_1}{.west}
    ;
    \path[bounce, bend right]
      \TBounce{a_1}{}{a_0}{.north west}
      \TBounce{b_1}{}{b_0}{.north west}
    ;
    \path[bounce, bend left]
      \TBounce{a_0}{}{a_1}{.south west}
      \TBounce{b_0}{}{b_1}{.south west}
    ;
    \path[bounce, bend left]
      \TBounce{c_0}{}{c_1}{.south west}
    ;
    \TState{a_1, b_0, ab_2, c_0}
  \end{tikzpicture}
  }
  \caption{
  \label{fig:ph-livelock}
    An example of PH with $2$ classes of priorities.
    Sorts are represented as labelled boxes and processes as circles with their identifier on the side.
    Actions of $\PHh^{(1)}$ are represented by thick arrows and actions of $\PHh^{(2)}$ are represented by thin arrows;
    the hit part of each action in drawn in plain line and the bounce part is in dotted line.
    Greyed processes stand for a possible state $s = \PHstate{a_1, b_0, c_0, ab_{10}}$.
  }
\end{figure}

@@

The construction of cooperation in PH allows to encode any Boolean function between cooperating processes \cite{PMR10-TCSB}.
Due to the introduction of priorities into the PH framework,
it is possible to build cooperations with no temporal shift by defining actions updating the cooperative sorts with the highest class of priority.
This allows to gain the same expressivity in PH than in Boolean networks, as stated in \pref{sec:dn}.
The aim of this paper is to allow the static analysis of the dynamics to be handled on PH models comprising such prioritised actions updating cooperative sorts.

@@

In the scope of this paper, we focus on a specific class of PH models
in which actions of $\PHh^{(1)}$ are called \emph{prioritized actions} and allow to model non-biological actions.
We consider in this section a PH model with $k$ classes of priorities: $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$, with $k \in \sNN$
and we define here the restrictions that lead to this class of models.

\pref{cr:bounded} states that the dynamics of the studied model $\PH$ contains no infinite sequence of prioritised actions.
As these actions can be considered as non-biological and therefore instantaneous, we thus prevent the existence of any Zeno-like behaviour
which would allow the play of an infinite sequence of prioritised actions in “zero time”.
As a consequence, the set $\restriction{\Sce}{1}$ is finite.

\begin{condition}[Bounded termination]
\label{cr:bounded}
  The dynamics of $\restriction{\PH}{1}$ contains no cycles:
  $\exists N \in \sN, \forall s \in \PHl, \forall \delta \in \restriction{\Sce}{1}(s), |\delta| \leq N$.
\end{condition}

@@

Our second condition (\pref{cr:compcs}) imposes that any sort in the Process
Hitting is either a well-formed component (\pref{def:component}) or a well-formed cooperative sort
(\pref{def:cs}).
The former is a sort that is hit only by actions with priority greater than $2$.
The latter is a sort that is hit only by actions with priority $1$ and which
always converges to the same process with respect to the state of the components
that have an influence on it (\pref{def:conn}).
We note $\components\subset\Sigma$ the set of well-formed components,
and $\cs\subset\Sigma$ the set of well-formed cooperative sorts.

@@

\begin{definition}[Components influence $(\compin: \PHs\to\powerset(\components))$]
\label{def:conn}
Given a sort $a$, $\compin(a)\DEF \conn(a)\cap\components$ where
$\conn(a) \subset \PHs$ is the smallest set of sorts satisfying the following properties:
\begin{align*}
a\in\conn(a) &
\\
\forall h\in  \PHh^{(1)},
	\Sigma(\target{h})\in\conn(a) & \Longrightarrow \Sigma(\hitter{h})\in\conn(a)
\end{align*}
\end{definition}

@@

We note that a sort $a$ that is hit by no action (\ie $\forall h \in \PHh, \PHsort(\target{h}) \neq a$)
does comply with both definitions of well-formed components and well-formed cooperative sorts.
In this case, such a sort can be arbitrarily affected to either of sets.
Such case may occur for “input” sorts (which state is initially fixed and do not evolve)
or some special cases of cooperative sorts produces by the translations of \pref{sec:encodings}.

@@

\begin{example}
  The PH in \pref{fig:ph-livelock} contains three components ($a$, $b$ and $c$) and a cooperative sort ($ab$, which models cooperation between sorts $a$ and $b$).
\end{example}

### 2014-03-28 parts/example.tex
global proofread

In this section, we illustrate the flattening method presented in \pref{ssec:flattening}
on a model of metazoan segmentation established in silico in \cite{MSB:MSB4100192}
in a differential equations framework.
It is composed of a wavefront gene $f$ that activates the gap-gene $a$ whose products are responsible or stripes.
Gene $f$ also activates a gene $c$ whose products repress the gene $a$.
The auto-inhibition of $c$ generalises a chain of repressors on $a$.
The auto-inhibition of $f$, which terminates the stripes formation in the original model,
have been removed in order to focus on the stationary dynamics of the model.

@@

    \path (1.8, 1.5) edge[superprio,coopupdate] (3.2, 2);
    \path (0.8, 4.5) edge[superprio,coopupdate] (3.2, 3);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-php}
    An example of PH $\oPH$ with $3$ classes of priorities.
    Actions of $\ov{\PHh}^{(3)}$ (lowest priority) are in thin lines,
    actions of $\ov{\PHh}^{(2)}$ (intermediate priority) are in thick lines,
    and actions of $\ov{\PHh}^{(1)}$ (highest priority), whose only purpose is to update the cooperative sort $fc$,
    are not represented but modeled by the double dashed arrows.
  }
\end{figure}

@@

\todo{Results on reachability?}

@@

    \path (-0.8, 1.5) edge[prio,coopupdate] (-2.2, 1.5);
    \path (4.3, 4.5) edge[prio,coopupdate] (-2.2, 2.5);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-ph}
    An example of PH $\PH$ with $2$ classes of priorities.
    Actions of $\PHh^{(2)}$ (unprioritized) are in thin lines,
    and actions of $\PHh^{(1)}$ (prioritized) which are used to update cooperative sorts,
    are abstracted by thick dashed arrows.
  }
\end{figure}

>>>>

In this section, we illustrate the flattening method presented in \pref{ssec:flattening}
on a model of metazoan segmentation inspired from a first translation to Process Hitting given in \cite{PMR10-TCSB}.
This model was originally established in silico in \cite{MSB:MSB4100192} in a differential equations framework.
It is composed of a wavefront gene $f$ that activates the gap-gene $a$ whose products are responsible for stripes formation.
Gene $f$ also activates a gene $c$ whose products represses the gene $a$.
The auto-inhibition of $c$ generalises a chain of repressors on $a$.
The auto-inhibition of $f$, which terminates the stripes formation in the original model,
has been removed in order to focus on the stationary dynamics of the model.

@@

    \path (1.8, 1.5) edge[superprio,coopupdate] (3.2, 2);
    \path (0.8, 4.5) edge[superprio,coopupdate] (3.2, 3);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-php}
    An example of PH $\oPH$ with $3$ classes of priorities,
    modelling the process of metazoan segmentation.
    Actions of $\ov{\PHh}^{(3)}$ (lowest priority) are in thin lines,
    actions of $\ov{\PHh}^{(2)}$ (intermediate priority) are in thick lines,
    and actions of $\ov{\PHh}^{(1)}$ (highest priority), whose only purpose is to update the cooperative sort $fc$,
    are not represented but modelled by the double dashed arrows.
  }
\end{figure}

@@

@@

    \path (-0.8, 1.5) edge[prio,coopupdate] (-2.2, 1.5);
    \path (4.3, 4.5) edge[prio,coopupdate] (-2.2, 2.5);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-ph}
    An example of PH $\PH$ with $2$ classes of priorities,
    which is the flattening of the PH in \pref{fig:metazoan-php}.
    Actions of $\PHh^{(2)}$ (unprioritised) are in thin lines,
    and actions of $\PHh^{(1)}$ (prioritised) which are used to update cooperative sorts,
    are abstracted by thick dashed arrows.
  }
\end{figure}

### 2014-03-28 parts/intro.tex
global proofread

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Usual formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Z$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Z$, without being itself changed.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in its multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more that four levels.

Besides, one further objective of our work is now %to be more accurate in the description of the studied systems.
to improve the accuracy of the description of the studied systems dynamics.
The idea for this is to introduce timing features into models:
we are interested in taking into account some knowledge about the relative length of some phenomena as it is a way to refute some models (or parameters) that are inconsistent with the observed dynamic behaviours.
In this paper, we are dealing with these timing properties through priorities,
that are based on the simple founding idea that prioritised actions have to be processed before the other ones.
Indeed, due to the Process Hitting framework restrictions, bimolecular reactions are not immediately available, but one can simulate them with an encoding called “cooperation”.
That encoding however introduces extra reactions, and this is where the priorities become useful, if not necessary.
The extra reactions can be given “infinite speed” (high priority) so that they do not affect the behaviour of “normal” (low priority) reactions, including the bimolecular ones.

@@

\paragraph{Contributions}
Since our formalism (the Process Hitting) can be considered as a subset of Calculus of Communicating Systems, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concepts of priority that allow for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refers to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a two-level static priority structure, some of them being designated as “prioritised” and others as “unprioritised”.

@@

Whereas the analysis presented in \pref{sec:sa} consider a particular class of
Process Hitting with two classes of priority,
we provide in \pref{sec:encodings} formal encodings of Asynchronous Discrete Networks and general
Process Hitting models with priorities.
Those encodings result in Process Hitting fulfilling the conditions for our
static analysis and show the broad applicability of our method.

>>>>

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Usual formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being itself changed.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in their multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more that four levels.

Besides, one further objective of our work is now %to be more accurate in the description of the studied systems.
to improve the accuracy of the description of the studied systems dynamics.
The idea for this is to introduce timing features into models:
we are interested in taking into account some knowledge about the relative length of some phenomena as it is a way to refute some models (or parameters) that are inconsistent with the observed dynamic behaviours.
In this paper, we are dealing with these timing properties through priorities,
that are based on the simple founding idea that actions with higher priority have to be processed before the ones with lower priority.
Furthermore, due to the Process Hitting framework restrictions, multimolecular reactions are not immediately available, but one can simulate them with an encoding called “cooperation”.
That encoding however introduces extra reactions, and this is where the priorities become useful, if not necessary.
The extra reactions can be given “infinite speed” (highest priority) so that they do not affect the behaviour of “normal” (lower priority) reactions, including the multimolecular ones.

@@

\paragraph{Contributions}
Since our formalism (the Process Hitting) can be considered as a subset of Calculus of Communicating Systems, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concept of priority that allows for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refers to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a static priority structure.
The actions modelling biological processes are divided in several classes of priorities which stand for the different reaction rates or relative preemptions.
Furthermore, the highest class of priority is used to model particular nonbiological actions, designated as “prioritised”,
that are used to encode Boolean functions inside the model.

@@

Whereas the analysis presented in \pref{sec:sa} considers a particular class of
Process Hitting with two classes of priority,
we provide in \pref{sec:encodings} formal encodings of Asynchronous Discrete Networks and general
Process Hitting models with priorities.
Those encodings result in a Process Hitting fulfilling the conditions for our
static analysis and show the broad applicability of our method.

### 2014-03-28 parts/models-ph-dn.tex
debut sequencial reachability
global proofread

We exhibit an encoding of Asynchronous Discrete Networks (ADN) with the Process
Hitting using two classes of priorities, and prove a weak bisimulation relation.

@@

\pref{th:bisimDN} states the weak bisimulation relation between an ADN and its encoding in
PH with $2$ classes of priorities.
Intuitively, actions updating cooperative sorts being prioritised, actions updating component
sorts follow strictly the possible transitions of the ADN.

>>>>

We exhibit an encoding of Asynchronous Discrete Networks (ADN) with the Process
Hitting using two classes of priorities
that respects all Conditions of \pref{ssec:hypothesis},
and prove a weak bisimulation relation.

@@

By construction, the cooperative sorts are well-formed, hence $\toPH\DNdef$
satisfies the Conditions of \pref{ssec:hypothesis}.
\pref{th:bisimDN} states the weak bisimulation relation between an ADN and its encoding in
PH with $2$ classes of priorities.
Intuitively, actions updating cooperative sorts being prioritised, actions updating component
sorts follow strictly the possible transitions of the ADN.

### 2014-03-28 parts/models-ph-php.tex
global proofread

\todo{Prove that $\PHflat(\oPH)$ respects all Conditions!}

@@

We showed in this subsection that it is possible to model a PH with $k$ classes of priorities as a PH
with $2$ classes of priorities with the same dynamics that respects all Conditions of \pref{ssec:hypothesis}.
We note however that the translation given in this section may not be minimal in terms of number of actions
or cooperative sorts.
We give here several ways to simplify the resulting model, but we do not give the proofs.

>>>>

@@

We showed in this subsection that it is possible to model a PH with $k$ classes of priorities as a PH
with $2$ classes of priorities and the same dynamics.
Furthermore, by construction, the cooperative sorts are well-formed, and $\PHflat(\oPH)$
thus satisfies the Conditions of \pref{ssec:hypothesis}.
We note however that the translation given in this section may not be minimal in terms of number of actions
or cooperative sorts.
We give here several ways to simplify the resulting model, but we do not give the proofs.
